Semi-permeable vesicles: mathematical formulation, numerics and physics

### 2020-03-31 vesicle_code/docs/permeableVes/main.tex
working on quiescent flow example

\usepackage{amsmath,amsfonts,bm, color}
\usepackage{graphicx,epsfig}
\usepackage{todonotes}
%Dfig puts a figure of specified size at the place indicated.
\newcommand{\Dfig}[2]{\epsfig{figure=#1,width=#2cm}}

@@

\section{Problem formulation} \label{sc:formulate}
Consider a two-dimensional semi-permeable vesicle suspended in an
unbounded viscous fluid domain, subjected to an imposed flow
$\uu_\infty(\xx)$, for any $\xx \in \RR^2$.  Assume that the interior
and exterior fluids have the same viscosity $\mu$. Let $\xx$ be the
position of the interface $\gamma$, $\uu$ be the fluid velocity, and
$p$ be the pressure. In the vanishing Reynolds number limit, the
governing equations for the ambient fluid are
%
\begin{subequations}
\begin{alignat}{3}
  \label{eqn:governing}
  -\nabla p + \mu \triangle \uu &= 0, 
  &&\xx \in \RR^2 \setminus \gamma, \\
% 
  \nabla \cdot \uu &= 0,  &&\xx \in \RR^2 \setminus \gamma, \\
%
  \uu(\xx) &\rightarrow \uu_\infty(\xx),
    \qquad &&\norm{\xx} \rightarrow \infty.
\end{alignat}
\end{subequations}%
%
Letting $\nn$ be the outward normal to $\gamma$, $\jump{\cdot}$ denote
the jump across the interface, and $T$ denote the hydrodynamic stress
tensor, the boundary conditions along the vesicle are
\begin{align}
  \label{eqn:jump1}
  \jump{T\cdot\nn} &= \ff_\text{mem}, \\
  \label{eqn:jump2}
  \jump{\uu} &= 0,
\end{align}
where $\ff_\text{mem}$ is the standard merrane force that includes a
bending and a tension term. In particular, $\ff_\text{mem} =
\ff_\text{ben} + \ff_\text{ten}$, where
\begin{align}
  \label{eqn:merraneForces}
  \ff_\text{ben} = -\xx_{ssss}, \qquad
  \ff_\text{ten} = (\sigma \xx_s)_s.
\end{align}
Since the bending stiffness only sets the time scale, we let $k_b = 1$
throughout this paper. The Lagrange multiplier $\sigma$ is used to
satisfy the inextensibility condition.  Finally, the tension of the
vesicle is
\begin{align}
  \Lambda = \sigma + 1.5 \kappa^2,
\end{align}
where $\kappa$ is the membrane curvature.

The kinematic condition is modified to account for the fact that the
fluid can move through the merrane. A standard choice appears
is~\cite{}
\begin{align}
  \uu - \dot{\xx} = - \beta (\ff_\text{mem} \cdot \nn) \nn, \qquad
  \xx \in \gamma,
\end{align}
where $\beta$ is a semi-permeability rate.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Vesicle Area}
Since the vesicle membrane is semi-permeable, the vesicle's area
fluctuates.  Parameterizing the vesicle with $\xx$, its area is
\begin{align}
  A(t) = \frac{1}{2}\int_{\gamma} (\xx \cdot \nn)\, ds .
\end{align}
The time derivative of $A(t)$ includes terms due to the time derivative
of $\xx \cdot \nn$, and a term due to interface
stretching~\cite{lai-tse-hua2008}. However, this second term is zero
because of the local inextensibility condition. Therefore, 
\begin{align}
  \dot{A}(t) &= \frac{1}{2} \int_{\gamma} \pderiv{}{t} 
    (\xx \cdot \nn) \, ds = 
  \frac{1}{2} \int_{\gamma} (\dot{\xx} \cdot \nn)\, ds  + 
  \frac{1}{2} \int_{\gamma} (\xx \cdot \dot{\nn})\, ds.
\end{align}
The second integrand can be computed as
\begin{align}
  \int_\gamma (\xx \cdot \dot{\nn})\, ds 
    = -\int_\gamma \left(\xx^\perp \cdot \dot{\xx}_s\right)\, ds 
    = \int_\gamma \left(\xx_s^\perp \cdot \dot{\xx}\right)\, ds 
    = \int_\gamma (\dot{\xx} \cdot \nn)\, ds.
\end{align}
Therefore,
\begin{align}
  \dot{A}(t) &= \int_\gamma (\dot{\xx} \cdot \nn)\, ds 
  = \int_\gamma \left(\beta (\ff \cdot \nn)\nn + \SS[\ff]\right) 
      \cdot \nn\, ds 
  = \beta \int_\gamma (\ff \cdot \nn)\, ds,
\end{align}
where the last equality comes follows from the divergence theorem.
Finally, using integration by parts and the Frenet-Serret identities, we
have
\begin{align}
  \dot{A}(t) &= \beta \int_\gamma \left( -\xx_{ssss} + 
    (\sigma \xx_s)_s \right) \cdot \nn \, ds \\
  &= \beta \int_\gamma \left(-\xx_{sss} + \sigma \xx_s 
    \right) \cdot \kappa \xx_s \, ds \\
  &= \beta \int_\gamma \left(\xx_{ss} \cdot 
    (\kappa \xx_s)_s + \kappa \sigma \right) \, ds \\
  &= \beta \int_\gamma \left(\kappa (\xx_{ss} \cdot \xx_{ss}) + 
    \kappa_s (\xx_{ss} \cdot \xx_s) + \kappa \sigma \right) 
    \, ds \\
  &= \beta \int_\gamma \left(\kappa^3 + \kappa \sigma \right) 
    \, ds.
\end{align}
Finally, using the relationship between the Lagrange multiplier
$\sigma$, and the tension $\Lambda$, the area of a semi-permeable
vesicle is governed by
\begin{align}
  \dot{A}(t) = \beta \int_\gamma \left(
    -\frac{\kappa^3}{2} + \kappa \Lambda \right) \, ds.
\end{align}

@@

\subsection{Discretization in Time}
By using an integral equation formulation, the only unknowns are the
vesicle position and tension, and the flow in the bulk does not have to
be computed. However, the high-order derivatives in the bending and
tension forces~\eqref{eqn:merraneForces} result in stiffness if an
explicit time stepping method is applied to
equation~\eqref{eqn:vesVelocity}. We apply a similar semi-implicit time
stepping method to our previous works~\cite{vee-gue-zor-bir2009,
qua-bir2014}. First, given a vesicle $\gamma$, with a boundary
parameterized as $\xx(s)$, we introduce notation for all the required
differential and projection operators
\begin{align}
  B[\xx]\ff &= -\frac{d^4}{ds^4} \ff, \\
  T[\xx]\sigma &= (\sigma \xx_s)_s, \\
  D[\xx]\ff &= \xx_s \cdot \ff_s, \\
  P[\xx]\ff &= (\ff \cdot \nn) \nn. 
\end{align}
Here, the normal vector $\nn$ is the outward unit normal of the
$\gamma$, and $d/ds$ is its arclength derivative. To simplify notation,
we write $B\ff$ to mean $B[\xx]\ff$, and use similar notation for the
other operators. Then, the dynamics of the vesicle are governed by
\begin{align}
  \frac{d\xx}{dt} &= \uu_{\infty}(\xx) + 
  \beta P(B\xx + T\sigma) + \SS(B\xx + T\sigma), \\
  D \frac{d\xx}{dt} &= 0.
\end{align}
By writing $B^N$ to mean the bending operator due to the vesicle
configuration at time $t^N$, and similar notation for the other
operators, we address the stiffness by introducing the semi-implicit
time stepping method
\begin{align}  
  \frac{\xx^{N+1} - \xx^N}{\Delta t} = \uu_\infty(\xx^N) 
  + \beta P^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right) 
  + \SS^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right).
\end{align}
Finally, the inextensibility condition is applied by requiring that 
\begin{align}
  D^N\xx^{N+1} = 1.
\end{align}

@@

We consider a single semi-permeable vesicle in a quiescent flow. For
impermeable vesicles, the vesicle relaxes to a shape with constant
negative tension~\cite{kra-win-sei-lip1996}.

@@

\section{Including a concentration field}
Based on the work of Yao and Mori~\cite{yao-mor2017}, we introduce a
chemical concentration field $c$.  The concentration field is governed
by the advection-diffusion equation
\begin{align}
  \pderiv{c}{t} + \nabla \cdot (\uu c) = D\Delta c.
\end{align}
If the concentration is diffusion dominated (zero Peclet number), then
the concentration satisfies
\begin{align}
  \Delta c = 0, \quad \xx \in \RR^2.
\end{align}
The boundary conditions of the concentration field on the vesicle
$\gamma$ are
\begin{align}
  (\uu c - D\nabla c) \cdot \nn = c \pderiv{\xx}{t} \cdot \nn + 
    j_c + j_p,
\end{align}
where $j_c$ and $j_p$ are the transmerrane chemical flux due to passive
transport and active transport (pumps), respectively.  In the zero
Peclet number limit, the absence of an active transport mechanism, and
the choice for the passive flux, the boundary condition of the
concentration field is
\begin{align}
  -D\pderiv{c}{\nn} = c \pderiv{\xx}{t} + 
    k_c \jump{c}, \quad \xx \in \gamma.
\end{align}
To couple the vesicle dynamics to the concentration, there is a
difference between the fluid velocity and vesicle velocity
\begin{align}
  \uu - \pderiv{\xx}{t} = j_w \nn,
\end{align}
where $j_w$ is the water flux through the vesicle.  This is the
osmophoretic flow.  In~\cite{yao-mor2017}, this flux is
\begin{align}
  j_w = -k_w(RT\jump{c} + \ff_\text{mem} \cdot \nn).
\end{align}
In our current implementation, we drop the $RT\jump{c}$ term, and
let $\beta = k_w$.

>>>>

%\usepackage{showkeys}
\usepackage{amsmath,amsfonts,bm}
\usepackage{graphicx,epsfig}
\usepackage{todonotes}
%Dfig puts a figure of specified size at the place indicated.
\newcommand{\Dfig}[2]{\epsfig{figure=#1,width=#2cm}}

@@

\section{Problem formulation} \label{sc:formulate}
Consider a two-dimensional semi-permeable vesicle suspended in an
unbounded viscous fluid domain, subjected to an imposed flow
$\uu_\infty(\xx)$, for any $\xx \in \RR^2$.  Assume that the interior
and exterior fluids have the same viscosity $\mu$. Let $\xx$ be the
position of the interface $\gamma$, $\uu$ be the fluid velocity, and
$p$ be the pressure. In the vanishing Reynolds number limit, the
governing equations for the ambient fluid are
%
\begin{subequations}
\begin{alignat}{3}
  \label{eqn:governing}
  -\nabla p + \mu \triangle \uu &= 0, 
  &&\xx \in \RR^2 \setminus \gamma, \\
% 
  \nabla \cdot \uu &= 0,  &&\xx \in \RR^2 \setminus \gamma, \\
%
  \uu(\xx) &\rightarrow \uu_\infty(\xx),
    \qquad &&\norm{\xx} \rightarrow \infty.
\end{alignat}
\end{subequations}%
%
Letting $\nn$ be the outward normal to $\gamma$, $\jump{\cdot}$ denote
the jump across the interface, and $T$ denote the hydrodynamic stress
tensor, the boundary conditions along the vesicle are
\begin{align}
  \label{eqn:jump1}
  \jump{T\cdot\nn} &= \ff_\text{mem}, \\
  \label{eqn:jump2}
  \jump{\uu} &= 0,
\end{align}
where $\nn$ points outward to the vesicle, and $\ff_\text{mem}$ is the
standard membrane force that includes a bending and a tension term. In
particular, $\ff_\text{mem} = \ff_\text{ben} + \ff_\text{ten}$, where
\begin{align}
  \label{eqn:membraneForces}
  \ff_\text{ben} = -\xx_{ssss}, \qquad
  \ff_\text{ten} = (\sigma \xx_s)_s.
\end{align}
Since the bending stiffness only sets the time scale, we let $k_b = 1$
throughout this paper. The Lagrange multiplier $\sigma$ is used to
satisfy the inextensibility condition.  Finally, the tension of the
vesicle is
\begin{align}
  \Lambda = \sigma + 1.5 \kappa^2,
\end{align}
where $\kappa$ is the membrane curvature.

The kinematic condition is modified to account for the fact that the
fluid can move through the membrane. A standard choice appears
is~\cite{}
\begin{align}
  \uu - \dot{\xx} = - \beta (\ff_\text{mem} \cdot \nn) \nn, \qquad
  \xx \in \gamma,
\end{align}
where $\beta$ is a semi-permeability rate.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Vesicle Area}
Since the vesicle membrane is semi-permeable, the vesicle's area
fluctuates.  Parameterizing the vesicle with $\xx$, its area is
\begin{align}
  A(t) = \frac{1}{2}\int_{\gamma} (\xx \cdot \nn)\, ds .
\end{align}
The time derivative of $A(t)$ includes terms due to the time derivative
of $\xx \cdot \nn$, and a term due to interface
stretching~\cite{lai-tse-hua2008}. However, this second term is zero
because of the local inextensibility condition. Therefore, 
\begin{align}
  \dot{A}(t) &= \frac{1}{2} \int_{\gamma} \pderiv{}{t} 
    (\xx \cdot \nn) \, ds = 
  \frac{1}{2} \int_{\gamma} (\dot{\xx} \cdot \nn)\, ds  + 
  \frac{1}{2} \int_{\gamma} (\xx \cdot \dot{\nn})\, ds.
\end{align}
The second integrand can be computed as
\begin{align}
  \int_\gamma (\xx \cdot \dot{\nn})\, ds 
    = -\int_\gamma \left(\xx^\perp \cdot \dot{\xx}_s\right)\, ds 
    = \int_\gamma \left(\xx_s^\perp \cdot \dot{\xx}\right)\, ds 
    = \int_\gamma (\dot{\xx} \cdot \nn)\, ds.
\end{align}
Therefore,
\begin{align}
  \dot{A}(t) &= \int_\gamma (\dot{\xx} \cdot \nn)\, ds 
  = \int_\gamma \left(\beta (\ff \cdot \nn)\nn + \SS[\ff]\right) 
      \cdot \nn\, ds 
  = \beta \int_\gamma (\ff \cdot \nn)\, ds,
\end{align}
where the last equality comes follows from the divergence theorem.
Finally, using integration by parts and the Frenet-Serret identities, we
have
\begin{align}
  \dot{A}(t) &= \beta \int_\gamma \left( -\xx_{ssss} + 
    (\sigma \xx_s)_s \right) \cdot \nn \, ds \\
  &= \beta \int_\gamma \left(-\xx_{sss} + \sigma \xx_s 
    \right) \cdot \kappa \xx_s \, ds \\
  &= \beta \int_\gamma \left(\xx_{ss} \cdot 
    (\kappa \xx_s)_s + \kappa \sigma \right) \, ds \\
  &= \beta \int_\gamma \left(\kappa (\xx_{ss} \cdot \xx_{ss}) + 
    \kappa_s (\xx_{ss} \cdot \xx_s) + \kappa \sigma \right) 
    \, ds \\
  &= \beta \int_\gamma \left(\kappa^3 + \kappa \sigma \right) 
    \, ds.
\end{align}
Finally, using the relationship between the Lagrange multiplier
$\sigma$, and the tension $\Lambda$, the area of a semi-permeable
vesicle is governed by
\begin{align}
  \dot{A}(t) = \beta \int_\gamma \left(
    -\frac{\kappa^3}{2} + \kappa \Lambda \right) \, ds.
  \label{eqn:areaROC}
\end{align}

@@

\subsection{Discretization in Time}
By using an integral equation formulation, the only unknowns are the
vesicle position and tension, and the flow in the bulk does not have to
be computed. However, the high-order derivatives in the bending and
tension forces~\eqref{eqn:membraneForces} result in stiffness if an
explicit time stepping method is applied to
equation~\eqref{eqn:vesVelocity}. We apply a similar semi-implicit time
stepping method to our previous works~\cite{vee-gue-zor-bir2009,
qua-bir2014}. First, given a vesicle $\gamma$, with a boundary
parameterized as $\xx(s)$, we introduce notation for all the required
differential and projection operators
\begin{align}
  B[\xx]\ff &= -\frac{d^4}{ds^4} \ff, \\
  T[\xx]\sigma &= (\sigma \xx_s)_s, \\
  D[\xx]\ff &= \xx_s \cdot \ff_s, \\
  P[\xx]\ff &= (\ff \cdot \nn) \nn. 
\end{align}
Here, the normal vector $\nn$ is the outward unit normal of $\gamma$,
and $d/ds$ is its arclength derivative. To simplify notation, we write
$B\ff$ to mean $B[\xx]\ff$, and use similar notation for the other
operators. Then, the dynamics of the vesicle are governed by
\begin{align}
  \frac{d\xx}{dt} &= \uu_{\infty}(\xx) + 
  \beta P(B\xx + T\sigma) + \SS(B\xx + T\sigma), \\
  D \frac{d\xx}{dt} &= 0.
\end{align}
By writing $B^N$ to mean the bending operator due to the vesicle
configuration at time $t^N$, and similar notation for the other
operators, we address the stiffness by introducing the semi-implicit
time stepping method
\begin{align}  
  \frac{\xx^{N+1} - \xx^N}{\Delta t} = \uu_\infty(\xx^N) 
  + \beta P^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right) 
  + \SS^N\left(B^N\xx^{N+1} + T^N\sigma^{N+1}\right).
\end{align}
Finally, the inextensibility condition is applied by requiring that 
\begin{align}
  D^N\xx^{N+1} = 1.
\end{align}

@@

We consider a single semi-permeable vesicle in a quiescent flow. For
impermeable vesicles, the vesicle relaxes to a shape where the tension
$\Lambda$ is constant and negative~\cite{kra-win-sei-lip1996}. In the
semi-permeable case, the area of the vesicle is governed by
equation~\eqref{eqn:areaROC}, where $\beta > 0$. We expect the
semi-permeable vesicle to reach a steady-state shape, meaning that in
the limit,
\begin{align}
  \int_{\gamma} \left(-\frac{\kappa^3}{2} + \kappa \Lambda \right) \, ds
  \rightarrow 0.
\end{align}
Based on numerical observations, the steady-state shape is a circular
vesicle with constant tension. Since the tension and curvature are
constant, we can solve for the curvature in terms of the tension $\kappa
= \sqrt{2\Lambda}$. Therefore, the steady state shape is a circle of
radius $r = 1/\sqrt{2\Lambda}$, and therefore the length is
\begin{align}
  L = \frac{\sqrt{2}\pi}{\sqrt{\Lambda}},
\end{align}
which is constant by the local inextensibility constraint. Therefore, we
can compute the final tension in terms of the length using the formula
\begin{align}
  \Lambda = \frac{2\pi^2}{L^2}.
  \label{eqn:finalTension}
\end{align}
Therefore, in contrast to a impermeable vesicle, the final tension is
positive rather than negative.

We start by simulating a star-shaped vesicle in a quiescent flow with
semi-permeability rate $\beta=1$. The vesicle shape at five time
steps is in Figure~\ref{fig:starShape}. To perform this simulation,
we use an adaptive second-order time stepping
method~\cite{qua-bir2016}. The times and reduced areas are
in the titles, and note that the shapes are not eqiuspaced in time.

\begin{figure}[htp]
  \begin{center}
    \input{figures/starRelaxation2.tikz}
    \input{figures/starRelaxation3.tikz}
    \input{figures/starRelaxation4.tikz}
    \input{figures/starRelaxation5.tikz}
    \input{figures/starRelaxation6.tikz}
  \end{center}
  \caption{\label{fig:starShape} A initially star-shaped
  semi-permeable vesicle in a shear quiescent flow with $\beta=1$. Note
  that the time steps are not equispaced.}
\end{figure}

\begin{figure}[htp]
  \input{figures/starAreaDeriv.tikz}
  \qquad
  \input{figures/starTension.tikz}
  \caption{\label{fig:starArea} (a) The area (red) and its derivative
  (blue) of an initially star-shaped vesicle as a function of time. The
  dots correspond to the time steps shown in Figure~\ref{fig:starShape}.
  (b) The tension of the vesicles illustrated in
  Figure~\ref{fig:starShape}.  The tension at the time horizon $T=1$ is
  also included. At this point, the vesicle is nearly circular and the
  tension agrees with the constant value from
  equation~\eqref{eqn:finalTension}. This dashed black line corresponds
  to this constant tension value.}
\end{figure}

@@

\section{Including a concentration field}
Based on the work of Yao and Mori~\cite{yao-mor2017}, we introduce a
chemical concentration field $c$.  The concentration field is governed
by the advection-diffusion equation
\begin{align}
  \pderiv{c}{t} + \nabla \cdot (\uu c) = D\Delta c.
\end{align}
If the concentration is diffusion dominated (zero Peclet number), then
the concentration satisfies
\begin{align}
  \Delta c = 0, \quad \xx \in \RR^2.
\end{align}
The boundary conditions of the concentration field on the vesicle
$\gamma$ are
\begin{align}
  (\uu c - D\nabla c) \cdot \nn = c \pderiv{\xx}{t} \cdot \nn + 
    j_c + j_p,
\end{align}
where $j_c$ and $j_p$ are the transmembrane chemical flux due to passive
transport and active transport (pumps), respectively.  In the zero
Peclet number limit, the absence of an active transport mechanism, and
the choice for the passive flux, the boundary condition of the
concentration field is
\begin{align}
  -D\pderiv{c}{\nn} = c \pderiv{\xx}{t} + 
    k_c \jump{c}, \quad \xx \in \gamma.
\end{align}
To couple the vesicle dynamics to the concentration, there is a
difference between the fluid velocity and vesicle velocity
\begin{align}
  \uu - \pderiv{\xx}{t} = j_w \nn,
\end{align}
where $j_w$ is the water flux through the vesicle.  This is the
osmophoretic flow.  In~\cite{yao-mor2017}, this flux is
\begin{align}
  j_w = -k_w(RT\jump{c} + \ff_\text{mem} \cdot \nn).
\end{align}
In our current implementation, we drop the $RT\jump{c}$ term, and
let $\beta = k_w$.
